---
title: 'Script 2: Descriptive Analysis of Pathogen and AMR Data'
author: "Heather Amato"
date: "2025-06-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install packages - if you haven't installed these yet, uncomment this line to install
# install.packages(c("janitor","patchwork","pheatmap"))

# Check R version
#version$version.string
install.packages("janitor")
install.packages("pheatmap")
install.packages("patchwork")
# Load libraries
library(here)
library(tidyverse)
library(janitor)
library(pheatmap)
library(patchwork) # patchwork requires R ≥ 3.3.0.

here() # make sure you are working in your GitHub folder
```

### Load cleaned datasets
```{r load.data}
# Workshop Task: Add code here to read in the cleaned microbio data and the cleaned tac data
microbio_clean <- read_csv(here( , ))
tac_clean <- read_csv(here( , ))
```

## TAC Data Analysis

#### Categorize TAC targets
```{r tac.categorize}
# Create lookup table of pathogen categories based on target name
tac_categories <- tibble::tibble( # tibble:: requires the package tibble (automatically installs before using)
  # get unique target names in data
  Target = unique(tac_clean$Target), 
  
  # identify any samples that include the following targets and assign a new value for the Category variable
  Category = case_when( 
    grepl("Ascaris|Trichuris|Strongyloides|Ancylostoma|Necator", Target, ignore.case = TRUE) ~ "Helminth",
    grepl("Giardia|Cryptosporidium|Entamoeba|Blastocystis|Cyclospora|Dientamoeba", Target, ignore.case = TRUE) ~ "Parasite",
    grepl("Adeno|Rota|Noro|Astro|Sapovirus|Enterovirus|Hepatitis", Target, ignore.case = TRUE) ~ "Virus",
    grepl("Shigella|Salmonella|Campylo|E. coli|ETEC|EPEC|EAEC|STEC|Vibrio|Yersinia|Clostrid|Aeromonas|Helicobacter|Listeria|Bacteroides", 
          Target, ignore.case = TRUE) ~ "Bacteria",
    grepl("bla|mecA|tetA|sul1|qnrS", Target, ignore.case = TRUE) ~ "ARG",
    grepl("HF183|CrAssphage|Sketa22|BacHum|MS2|Spiked|Control|NTC", Target, ignore.case = TRUE) ~ "Control",
    TRUE ~ "Other"
  )
)

# Merge categories into the TAC data
tac_clean <- tac_clean %>%
  # left join merges the tac_categories (right) to tac_clean (left)
  left_join(., tac_categories, by = "Target") 
```

#### Calculate prevalence of each TAC target
```{r tac.prevalence}
# Create summarized data set of target prevalence and standard deviation
tac_prev <- tac_clean %>%
  group_by(Category, Target) %>% 
  summarize(prevalence = mean(detect, na.rm = TRUE), # remove NA values in calculations
            sd = sd(detect, na.rm = TRUE)) %>%
  arrange(desc(prevalence)) # show pathogen targets in order of highest to lowest prevalence

# Take a peak at the top 10 (and bottom 10) pathogens by prevalence
head(tac_prev, 10)
tail(tac_prev, 10)
```

#### Visualize prevalence of TAC pathogen targets
```{r plot.tac.prevalence}
# Start with prevalence dataframe you created above, then pipe it into your ggplot() code
tac_prev %>%
  filter(Category != "Control" & Category != "ARG") %>% # remove control samples & ARG targets
  ggplot(aes(x = reorder(Target, prevalence), # plot x (Target) in order of prevalence values
             y = prevalence, color = Category)) +  # color points by pathogen category
  geom_point() + 
  scale_color_manual(values = c("cadetblue3","purple4","yellowgreen","orange2")) + # one color for each pathogen category
  coord_flip() + # flip x and y axis so prevalence is on x axis
  labs(title = "Pathogen Prevalence (TAC)", x = "Target", y = "Prevalence") + 
  theme_bw(base_size = 11) + # use this for set themes - e.g. theme_minimal(), theme_classic(), etc.
  theme(axis.ticks.y = element_blank(), # customize theme (remove axis ticks, set axis label size)
        axis.text.y = element_text(size = 8, face = "italic"))

# Save plot
ggsave( here("figures_answers", "pathogen_prevalence_scatterplot.png"), dpi = 300, width = 7, height = 5, unit = "in")
```
#### Calculate prevalence of each TAC target by sample type
```{r tac.prevalence.type}
# Workshop Task: how would you modify the above code to create a new tac_prev_type dataset by category, target, and sample type?
tac_prev_type <- tac_clean %>%
  group_by(Category, Target) %>% 
  summarize(prevalence = mean(detect, na.rm = TRUE), # remove NA values in calculations
            sd = sd(detect, na.rm = TRUE)) %>%
  arrange(desc(prevalence)) # show pathogen targets in order of highest to lowest prevalence

# Workshop Task: ADD CODE HERE to take a peak at the top 10 (and bottom 10) pathogens by prevalence


```

#### Visualize prevalence of TAC targets by sample type
Now we're going to repeat the plot above, but group by pathogen category, target and sample type and build a multi-panel plot using facet_wrap(). facet_wrap() and facet_grid() allow you to split your single plot into multiple plots, or stratify, by a categorical/factor variable. For this plot, we will color the points by sample_type and facet by pathogen category.  
```{r plot.tac.prevalence.type}
# Workshop task: Practice commenting out one line at a time and rerun the code to see how the plot changes
tac_prev_type %>% 
  filter(Category != "Control" & Category != "ARG") %>% # remove control samples & ARG targets
  ggplot(aes(x = reorder(Target, prevalence), # plot x (Target) in order of prevalence values
             y = prevalence, color = sample_type)) +  
  geom_point() +
  facet_wrap(~Category, scales = "free_y") + # add a facet here to separate plots by target category
  coord_flip() + # flip x and y axis so barplot is sideways
  labs(title = "Pathogen Prevalence by Sample Type (TAC)", x = "Target", y = "Prevalence") +
  theme_bw(base_size = 11) + # use this for set themes - e.g. theme_minimal(), theme_classic(), etc.
  theme(axis.ticks.y = element_blank(), # customize theme (remove axis ticks)
        axis.text.y = element_text(size = 8, face = "italic"),
        legend.position = "bottom")

# Save plot
ggsave(here("figures_answers", "pathogen_prevalence_scatterplot_faceted.png"), dpi = 300, width = 7, height = 5, unit = "in")
```

#### Summarize pathogen burden per sample
```{r burden.tac}
tac_burden <- tac_clean %>%
  filter(Category != "Control") %>% 
  group_by(household_id, sample_type) %>% # for each household and sample type
  summarize(num_detected = sum(detect, na.rm = TRUE)) # sum total detects

# Workshop Task: ADD CODE HERE to print a summary() of the new variable of total number detected for each sample

```

### Visualize distribution of pathogen burden overall
```{r plot.tac.burden}
# Workshop Task: use ggplot to plot a histogram - complete the code below to plot the tac burden variable you created in the previous code chunk, color coded by sample type
tac_burden %>%
  ggplot(aes(x = , fill = )) + 
  geom_histogram(binwidth = 1, alpha = 0.8) + # test out what happens when you change binwidth 
  labs(title = , fill = , x = , y = ) +
  theme_bw() 

# Save plot
ggsave( here("figures_answers", "pathogen_burden_histogram.png"), dpi = 300, width = 7, height = 5, unit = "in")
```

#### Visualize pathogens per sample by target category and sample type 
Here you will summarize the tac_burden dataframe further to estimate the mean number of pathogens detected per sample type and pathogen category.  
```{r pathogen.heatmap}
# Create dataframe for plot
tac_burden_cat <- tac_clean %>%
  filter(Category != "Control") %>% 
  # first get num_detects for each household and sample type as you did above
  group_by(Category, household_id, sample_type) %>% 
  summarize(num_detected = sum(detect, na.rm = TRUE)) %>%  # sum total detects
  # now get the mean num_detected per sample
  group_by( sample_type, Category) %>% 
  summarize(mean_detected = mean(num_detected, na.rm = TRUE)) # mean total detects per sample by category

# Set Category variable as ordered factor (to specify in which order they will appear in the plot)
tac_burden_cat$Category <- factor(tac_burden_cat$Category, 
                                   levels = c("Other", "Helminth","Parasite","Virus","Bacteria","ARG"))

# Plot as a heatmap
tac_burden_cat %>% 
  ggplot( aes(x = sample_type, y = Category, fill = mean_detected)) +
    geom_tile(color = "white") +
    scale_fill_viridis_c(option = "C", name = "Mean # Pathogens\nper Sample") + # adding \n splits into two rows
    labs(title = "Pathogen Burden by Category and Sample Type",
         x = "Sample Type", y = "Category") +
    theme_minimal(base_size = 11) + # set font size for entire plot
    coord_fixed() + # forces x and y axes to be plotted in proportion (squares instead of rectangles)
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Save plot
ggsave( here("figures_answers", "pathogen_heatmap.png"), dpi = 300, width = 5, height = 7, unit = "in")
```


## Culture Data Analysis
#### Prevalence of bacterial isolates
```{r idexx.prevalence}
# Workshop Task: complete code below to use summarize all detect variables to get prevalence
microbio_clean %>%
  group_by(sample_type) %>% 
  summarize(
    ecoli_positive = ,
    ar_ecoli_positive = ,
    tc_positive = ,
    ar_tc_positive = ,
    esbl_ec_positive = 
  )
```


#### Visualize ESBL prevalence by sample type
Using the microbiology cleaned dataset, we will plot the overall prevalence of ESBL-*E. coli* by sample type.  
```{r esbl.plot}
# Workshop Task: practice uncommenting line by line and rerunning; alter the specifications vjust, size, and color in geom_text to see how it changes the labels
microbio_clean %>%
  group_by(sample_type) %>%
  summarize(esbl_prev = mean(esbl_ecoli_detect), 
            esbl_prev_sd = sd(esbl_ecoli_detect)) %>%
  ggplot(aes(x = sample_type, y = esbl_prev, fill = sample_type)) +
  geom_col() +
  geom_text(aes(label = paste0(round(esbl_prev,2), " (", round(esbl_prev_sd,2), ")")), # add labels
                vjust = -0.5, color = "black", size = 3.5) +
  labs(title = "ESBL Prevalence by Sample Type", y = "Prevalence", 
       x = "Sample Type", fill = "Sample Type") +
  theme_minimal(base_size = 11) 

# Save plot
ggsave( here("figures_answers", "ESBL_barchart.png"), dpi = 300, width = 5, height = 3, unit = "in")
```
#### Visualize ESBL concentrations by sample type
Using the microbiology cleaned dataset, we will plot the log-transformed CFU of ESBL-*E. coli* by sample type.  

```{r esbl.concentration}
# Workshop Task: Practice commenting out individual lines of this code to see how it changes the plot (for example, first run the code as is, then uncomment geom_jitter and comment geom_point, then rerun the plot)
microbio_clean %>% 
  ggplot(aes(x = sample_type, y = log_adj_esbl_ecoli, fill = sample_type)) +
  geom_boxplot(aes(fill = sample_type), alpha = 0.6, outliers = FALSE) +
  geom_point(aes(color = sample_type)) + # points plotted on grid lines
# geom_jitter(aes(color = sample_type), width = .1)  + # points jittered around grid lines
  labs(fill = "Sample Type", 
       x = "Sample Type", y = "log10(CFU)", 
       title = expression("ESBL-" * italic("E. coli") * " Concentrations by Sample Type")) +
 # guides(color = "none") + # removes legend for color = sample_type
  theme_bw() 
```



## Merge TAC and Microbio Datasets by Sample 
#### Join by household ID
Reshaping and merging data are key tasks in data processing for analysis. 
```{r join.data}
# Reshape data from long to wide (one row per sample)
tac_household <- tac_clean %>% 
  select(household_id, sample_type, Target, detect) %>% 
  pivot_wider(names_from = Target, values_from = detect) 

# Join tac_household (right) to microbio_clean (left) using a left_join
tac_wide <- left_join(microbio_clean, tac_household, by = c("household_id", "sample_type")) 
```

#### Concordance Between TAC and Culture for *E. coli*
Create a two-by-two table to check whether the culture results for *E. coli* match *E. coli* O157 detection from TAC results.  Theoretically, where there is *E. coli* there may be pathogenic *E. coli* O157, but there should not be *E. coli* O157 where there is no *E. coli* detected as a fecal indicator bacteria.  
```{r concordance.ec}
tac_wide %>%
  select(ec_detect, `E. coli O157`) %>% # add quotes around variable names now because of space in names
  tabyl(ec_detect, `E. coli O157`) %>% # generate 2x2 frequency table using binary detection variables
  adorn_percentages("row") %>% # add row percent
  adorn_pct_formatting(digits = 1) # round to one digit
```

#### Concordance Between TAC and Culture for ESBL (beta-lactamase resistance)
Repeat the process above, but now compare TAC results for beta-lactamase resistance gene blaCTX-M with culture-based detection of ESBL *E. coli*.  
```{r concordance.esbl}
# Workshop Task: repeat the same code above, but this time to compare ESBL-E. coli and blaCTX-M variables
tac_wide %>%
  select( ) %>% 
  tabyl( ) %>%
  adorn_percentages() %>%
  adorn_pct_formatting()
```

#### Plot heatmap of ARG detection by ESBL-*E. coli* detection
In theory, we would expect to see frequent detection of beta-lactamase gene, blaCTX-M, for ESBL-*E. coli* positive samples. Let's see if that's the case in our simulated dataset. We will plot a heatmap of ARG detection with different fill colors to indicate ESBL-positive samples from the culture results.
```{r arg.heatmap}
# Join micro and tac data, but keep in long format because we want to plot by Target and Category 
tac_join <- left_join(microbio_clean, tac_clean, by = c("household_id", "sample_type"))

# Create dataset for plot
heatmap_dat <- tac_join %>%
  filter(Category == "ARG") %>% # select only ARG targets
  mutate(
    # specify the fill color to differ based on ESBL detection
    fill_color = case_when(
      detect == 1 & esbl_ecoli_detect == 1 ~ "Detected with ESBL", 
      detect == 1 & esbl_ecoli_detect == 0 ~ "Detected without ESBL",
      TRUE ~ "Not detected"))

heatmap_dat %>% # Pipe in dataset to ggplot()
  ggplot(aes(x = household_id, y = Target, fill = fill_color)) +
  geom_tile(color = "white") + # set background color
  scale_fill_manual( # set colors for ESBL+ samples, ESBL- samples, and no ARG detects
    values = c(
      "Detected with ESBL" = "purple",
      "Detected without ESBL" = "lightblue",
      "Not detected" = "white"), 
    name = "Detection") + # set label for color legend
  labs(title = "ARG Detections by Household",
       x = "Household Samples",
       y = "ARG Target") +
  facet_wrap(~sample_type, scales = "free_x", nrow = 3) + # stratify plot by sample type with each type in a different row
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank())

# save plot
ggsave( here("figures_answers", "ARG_ESBL_detects.png"), dpi = 300, width = 7, height = 5, unit = "in")
```

#### Plot Pathogen Detection by Soil Moisture (for Soil and Effluent)
We may be interested in how different sample parameters measured in the lab influence pathogen abundance. Here, we can plot moisture content against pathogen detection for each sample, and add a linear regression line to show the relationship between the two continuous variables. This may not be one of the main research questions, but could be useful to include in Supporting Information for your manuscript.  
```{r tac.moisture}
# Count number of detected pathogens per sample
pathogen_counts <- tac_clean %>%
  filter(sample_type != "produce") %>% # remove produce samples since those would not have moisture content data
  filter(Category %in% c("Bacteria", "Virus", "Parasite", "Helminth")) %>%  # optional: exclude controls/ARGs
  group_by(household_id, sample_type) %>%
  summarise(n_pathogens = sum(detect == 1, na.rm = TRUE)) %>%
  ungroup()

# Join with sample metadata that includes soil moisture
df_plot <- pathogen_counts %>%
  left_join(microbio_clean, by = c("household_id", "sample_type"))

# Plot
df_plot %>% 
  ggplot( aes(x = soil_moisture, y = n_pathogens, color = sample_type)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "black", linetype = "dashed") +
  labs(
    x = "Soil Moisture (%)",
    y = "Number of Pathogens Detected",
    title = "Pathogen Count vs. Soil Moisture",
    color = "Sample Type"
  ) +
  theme_minimal(base_size = 11) 

# save plot
ggsave( here("figures_answers", "pathogen_count_moisture.png"), dpi = 300, width = 7, height = 5, unit = "in")
```


## Save Outputs 
```{r save.results, eval=FALSE}
write_csv(tac_prev, here("outputs_answers", "tac_prevalence.csv")) # create "outputs" folder before saving
```

