---
title: 'Script 1: Data Processing (Answers Version)'
author: "Heather Amato"
date: "2025-06-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install packages
#install.packages(c("here","tidyverse","dplyr","readxl"))

#install.packages("remotes")                        # if not already installed
#remotes::install_github("jknappe/quantitray")      # installs the package directly from GitHub

# Load libraries
library(here)
library(tidyverse)
library(dplyr)
library(quantitray)
library(readxl)

# Check current directory & set new directory if needed
here()
```

#### Read data into RStudio
```{r read.dat}
# Read microbiological data
# (there are also packages you can install and load, like read_xl(), for reading .xlsx files)
microbio_df <- read_csv(here("simulated_data", "microbial_data_simulated.csv"))

# Folder path for TAC .csvs
tac_folder <- here("simulated_data", "simulated_cards")

# Read and combine all TAC .xlsx output files into tac_raw
tac_raw <- list.files(tac_folder, full.names = TRUE, pattern = "\\.xlsx$") %>%
  map_dfr(read_csv, .id = "source_file", show_col_types = FALSE)
```

## MICROBIAL DATA
#### Check micro (IDEXX) data format
```{r micro.check}
# Code provided to check structure
names(microbio_df) # variable names
summary(microbio_df) # summary of data
head(microbio_df, n=8) # view top 8 rows
unique(microbio_df$sample_id) %>% length() # number of samples
table(microbio_df$sample_type) # number of rows in dataset by type
```

#### Clean and process microbio dataset
```{r clean.micro}
microbio_clean <- microbio_df %>%
  mutate(
    household_id = str_extract(sample_id, "HH\\d+"), # extract household id from sample_id
    sample_type = str_extract(sample_id, "effluent|compost|produce"), # extract sample type from sample_id
    wet_soil_mass = wet_weight - plate_weight,
    dry_soil_mass = dry_weight - plate_weight,
    soil_moisture = (wet_soil_mass - dry_soil_mass) / dry_soil_mass *100 # estimate soil moisture content as percent
  ) %>%
  dplyr::select(
    -c(sample_id, plate_weight, wet_weight, dry_weight, wet_soil_mass, dry_soil_mass)
    )

# Check new variables
table(microbio_clean$sample_type)
summary(microbio_clean$soil_moisture)
```

#### Estimate Most Probable Number (MPN) from IDEXX data
```{r mpn}
microbio_clean <- microbio_clean %>%
  mutate(
    # Total coliform
    tc_mpn = quantify_mpn(total_coliform_large_cell_count, total_coliform_small_cell_count, method = "qt-2000"),
    tc_mpn_lo = quantify_95lo(total_coliform_large_cell_count, total_coliform_small_cell_count, method = "qt-2000"),
    tc_mpn_hi = quantify_95hi(total_coliform_large_cell_count, total_coliform_small_cell_count, method = "qt-2000"),
    # E. coli
    ecoli_mpn = quantify_mpn(e_coli_large_cell_count, e_coli_small_cell_count, method = "qt-2000"),
    ecoli_mpn_lo = quantify_95lo(e_coli_large_cell_count, e_coli_small_cell_count, method = "qt-2000"),
    ecoli_mpn_hi = quantify_95hi(e_coli_large_cell_count, e_coli_small_cell_count, method = "qt-2000"),
    # antibiotic-resistant TC
    ar_tc_mpn = quantify_mpn(ar_total_coliform_large_cell_count, ar_total_coliform_small_cell_count, method = "qt-2000"),
    ar_tc_mpn_lo = quantify_95lo(ar_total_coliform_large_cell_count, ar_total_coliform_small_cell_count, method = "qt-2000"),
    ar_tc_mpn_hi = quantify_95hi(ar_total_coliform_large_cell_count, ar_total_coliform_small_cell_count, method = "qt-2000"),
    # antibiotic-resistant E. coli
    ar_ecoli_mpn = quantify_mpn(ar_e_coli_large_cell_count, ar_e_coli_small_cell_count, method = "qt-2000"),
    ar_ecoli_mpn_lo = quantify_95lo(ar_e_coli_large_cell_count, ar_e_coli_small_cell_count, method = "qt-2000"),
    ar_ecoli_mpn_hi = quantify_95hi(ar_e_coli_large_cell_count, ar_e_coli_small_cell_count, method = "qt-2000")
  )
```

```{r mpn.check}
# Check dataset
microbio_clean %>% 
  select(household_id, sample_type, tc_mpn:ar_ecoli_mpn_hi) %>% 
  head(10)
```

```{r mpn.nafix}
# Replace NA values with 0s and create detection indicators
microbio_clean <- microbio_clean %>% 
  mutate(
    across(everything(), ~replace_na(.x, 0)), # replace NAs with 0s across all variables in dataframe
    tc_detect = if_else(tc_mpn > 0, 1, 0), # set detect = 1 if mpn > 0
    ar_tc_detect = if_else(ar_tc_mpn > 0, 1, 0),
    ec_detect = if_else(ecoli_mpn > 0, 1, 0),
    ar_ec_detect = if_else(ar_ecoli_mpn > 0, 1, 0)
  )
```

#### Transform ESBL CFU results
Plate count data for ESBL-E. coli should be adjusted for moisture content and sample volume (for effluent and compost samples only), and should be log-transformed to create a normal (bell-curve) distribution for statistical analyses.  
```{r plot.cfu}
hist(microbio_clean$esbl_e_coli_cfu) # check distribution with histogram
```

```{r adjust.esbl}
microbio_clean2 <- microbio_clean %>%
  mutate(moisture_fraction = soil_moisture / 100) %>%
  mutate(esbl_e_coli_cfu_adj = if_else(esbl_e_coli_cfu == 0, 0.5, esbl_e_coli_cfu),
         adjusted_esbl_cfu = case_when(
           sample_type %in% c("effluent", "compost") ~ esbl_e_coli_cfu_adj / (1 - moisture_fraction) / 2,
           sample_type == "produce" ~ esbl_e_coli_cfu_adj,
           TRUE ~ NA_real_ ),
         log_adjusted_esbl_cfu = log10(adjusted_esbl_cfu) + 1)

hist(microbio_clean2$adjusted_esbl_cfu)
hist(microbio_clean2$log_adjusted_esbl_cfu)
```

``` {r hist.esbl.samples}
# further investigate distribution with a histogram by sample type
ggplot(microbio_clean2, aes(x = log_adjusted_esbl_cfu, fill = sample_type)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  theme_minimal()
```

#### Transform MPN results
Next we will transform IDEXX MPN results. This process is similar to the above steps taken for ESBL E. coli, but now we will do this across multiple variables (each of the bacterial indicators).  
```{r adjust.mpn}
microbio_clean2 <- microbio_clean2 %>%
  mutate(across(c(tc_mpn, ar_tc_mpn, ecoli_mpn, ar_ecoli_mpn),
                ~ if_else(. == 0, 0.5, .),
                .names = "{.col}_adj")) %>%
  mutate(across(matches("_adj$"),
                ~ if_else(sample_type %in% c("compost", "effluent"), . / (1 - moisture_fraction) / 2, . / 2),
                .names = "adjusted_{.col}")) %>%
  mutate(across(matches("^adjusted_.*_adj$"), ~ log10(.) + 1, .names = "log_{.col}"))

# Check distributions before vs. after transforming
hist(microbio_clean2$tc_mpn) ; hist(microbio_clean2$ecoli_mpn)
hist(microbio_clean2$ar_tc_mpn) ; hist(microbio_clean2$ar_ecoli_mpn)
hist(microbio_clean2$log_adjusted_tc_mpn_adj);  hist(microbio_clean2$log_adjusted_ecoli_mpn_adj)
hist(microbio_clean2$log_adjusted_ar_tc_mpn_adj); hist(microbio_clean2$log_adjusted_ar_ecoli_mpn_adj)
```

## TAC DATA
#### Check TAC data format
```{r}
# Check TAC data
names(tac_raw)
summary(tac_raw)
head(tac_raw)

unique(tac_raw$Sample) %>% length() # number of unique samples
unique(tac_raw$Sample[grepl("NTC", tac_raw$Sample)]) %>% length() # number of non-template controls
unique(tac_raw$Sample[grepl("NTC", tac_raw$Sample) & tac_raw$Result == "Positive"]) %>% length() # N failed NTCs
```

#### Clean TAC data
```{r clean.tac}
# Workshop task: Clean TAC data and extract variables
tac_clean <- tac_raw %>%
  filter(!grepl("NTC", Sample, ignore.case = TRUE)) %>% # remove non-template controls

# extract household_id and sample_type and create binary detect variable here
  mutate(
    household_id = str_extract(Sample, "HH\\d+"),
    sample_type = str_extract(Sample, "effluent|compost|produce"),
    detect = if_else(Result == "Positive", 1, 0)
  )

View(tac_clean)
```

## Save final datasets
Select whatever variables you need for the analysis and save the cleaned datasets in a folder (clean_data) within the folder you created for the GitHub repository. Make sure you create the clean_data folder before running the code.  
```{r final.data}
microbio_clean3 <- microbio_clean2 %>%
  select(household_id, sample_type, soil_moisture, esbl_ec_cfu = esbl_e_coli_cfu,
         log_adj_esbl_ecoli = log_adjusted_esbl_cfu, esbl_ecoli_detect,
         log_adj_tc_mpn = log_adjusted_tc_mpn_adj, log_adj_ec_mpn = log_adjusted_ecoli_mpn_adj,
         log_adj_ar_tc_mpn = log_adjusted_ar_tc_mpn_adj, log_adj_ar_ec_mpn = log_adjusted_ar_ecoli_mpn_adj,
         tc_detect, ec_detect, ar_tc_detect, ar_ec_detect)

write_csv(microbio_clean3, here("clean_data", "microbial_data_cleaned.csv")) # clean_data is folder in current directory

write_csv(tac_clean %>%
            select(household_id, sample_type, Target, Well, Cq, Cq_sd = `Cq SD`, Result, detect),
          here("clean_data", "tac_data_cleaned.csv"))
```